---
title: "Merge Allen"
format: gfm
---

This is a simple script to download the Allen Institute Brain atlas (ccfv3) and merge smaller ROI togethers as a function of their parent branches. 


These are the packages you need to install and import before running the script.
```{r}
library(xml2)
library(nat)
library(oro.nifti)
library(abind)
```

Set the working directory to where you want to save the files.
```{r}
setwd('/project/4180000.36/merge_allen')
```

Download the annotation file from the Allen Institute website. This is a 3D array with the same size as the Allen Institute Brain atlas. The array contains the ROI ID for each voxel.
```{r}
#download allen map
file.address<-"http://download.alleninstitute.org/informatics-archive/current-release/mouse_ccf/annotation/ccf_2017/annotation_25.nrrd"
download.file(file.address,'annotation_25.nrrd')
annotation<-read.nrrd('annotation_25.nrrd')
lr.limit<-dim(annotation)[3]/2
```

Make a blank array with the same size as the annotation file. This will be used to store the new ROI ID.
```{r}
#make an empty array with the same size as the annotation
annotation.blank<-array(0,dim = dim(annotation))

#make a left and right hemisphere array
annotation.left<-array(0,dim=c(dim(annotation)[1:2],lr.limit))
annotation.right<-array(1,dim=c(dim(annotation)[1:2],lr.limit))
annotation.hemisphere<-abind(annotation.left,annotation.right,along = 3)
```

This is the most important part of the script. roi.branch is where you list the parent roi "ID" that you want to keep. All branches from the parents will be merged.
```{r}
#list of ROIs with branches to merge
roi.branch<-c(184, 500, 353, 329, 337, 345, 369, 361, 182305689, 378, 1057, 677, 247, 669, 31, 972, 44, 714, 95, 254, 22, 541, 922, 895, 1080, 822, 703, 672, 56, 803, 549, 1097, 313, 771, 354, 512)
roi.exclude<-c()
```

Download the roi list from the Allen API and put it into a dataframe. There is an option to exclude some ROIs if you don't want sub-branches to be merged with others. 
```{r}
xml.file<-'http://api.brain-map.org/api/v2/structure_graph_download/1.xml'


d<-read_xml(xml.file)

id <- xml_double(xml_find_all(d, ".//id"))
parent.id<-xml_double(xml_find_all(d, ".//parent-structure-id"))
acronym<-xml_text(xml_find_all(d, ".//acronym"))
name<-xml_text(xml_find_all(d, ".//name"))
hex<-xml_text(xml_find_all(d, ".//color-hex-triplet"))
level<-xml_double(xml_find_all(d, ".//st-level"))

dd<-data.frame(id,parent.id,acronym,name,hex,level)
#dd<-dd[-c(which(dd$id %in% roi.exclude)),]
```

This is the main loop. It will go through each parent ROI and merge all the branches together. 
```{r}
d.roi.branch.list<-c()

for(i in 1:length(roi.branch)){
  
roi.branch.list<-c()
  
d.id.select<-which(dd$id==roi.branch[i])
d.level.select<-dd$level[d.id.select]

print(paste('now doing ', dd$acronym[d.id.select], round(100*i/length(roi.branch),2), '% left', sep=''))

roi.branch.list<-roi.branch[i]

if(d.level.select<max(dd$level)){
for(j in (d.level.select+1): max(dd$level)){
  roi.branch.list<-unique(c(roi.branch.list,dd$id[which(dd$parent.id %in% roi.branch.list)]))
}}

d.roi.branch.list[i]<-paste(roi.branch.list,collapse = ',')

l.select<-which(annotation %in% roi.branch.list & annotation.hemisphere==0)
r.select<-which(annotation %in% roi.branch.list & annotation.hemisphere==1)
  
annotation.blank[l.select]<-i
annotation.blank[r.select]<-i+length(roi.branch)
}
```

```{r}
dd.select<-rbind(dd[dd$id %in% roi.branch,],dd[dd$id %in% roi.branch,])
dd.select$branches<-c(rep(d.roi.branch.list,2))
dd.select$hemisphere<-c(rep('left',length(roi.branch)),rep('right',length(roi.branch)))
dd.select$new.id<-1:(length(roi.branch)*2)


write.csv2(dd.select,'ABI_atlas_reduced.csv')

nii<-nifti(annotation.blank,pixdim=c(1, 0.025,0.025,0.025,1),datatype=8)
writeNIfTI(nii,'ABI_atlas_reduce',onefile=TRUE,gzipped=TRUE)
#save(list=c('annotation.blank'))
```


finally, we need to swap the image dimensions using FSL and add axis labels. I do this with AFNI. 
```{bash}
cd /project/4180000.36/merge_allen
fslswapdim ABI_atlas_reduce.nii.gz z -x -y swap.nii.gz
3dresample -prefix ABI_atlas_reduce_resample.nii.gz -inset swap.nii.gz
```
